Extracting Function Signature from R Program Traces

### 2020-05-12 paper/main.tex
possibly NA primitives

\lstset{language=R}
\definecolor{LightGray}{rgb}{.92,.92,.92}
\definecolor{Gray}{rgb}{.3,.3,.3}
\definecolor{DarkGray}{rgb}{.5,.5,.5}
\lstset{ %
  columns=flexible,
  captionpos=b,
  frame=single,
  framerule=0pt,
  tabsize=2,
  belowskip=0.5em,
  backgroundcolor=\color{LightGray},
  basicstyle=\small\ttfamily,
  emphstyle=,
  keywordstyle=,
  commentstyle=\color{Gray}\em,
  stringstyle=\color{Gray}
}
\lstdefinestyle{R}{ %
  language=R,
  morekeywords={assign, delayedAssign},
  deletekeywords={env, equal, c, runif, trace, args, exp, t, all},
  breaklines=true
}
\lstdefinestyle{Rin}{ %
  style=R,
  breaklines=false
}

@@

% %
% %
% \subsubsection{Cross-Validation Against Kaggle}

% Kaggle is a website offering an environment for hosting Jupyter notebooks, and importantly provides access to a large number of open-source data sets and R and Python code.
% This solved a major issue with R programs hosted on e.g. Github, where often the data is not made available.
% Our aim is to run R programs hosted on Kaggle to validate the type signatures we generated for popular R packages.

% \AT{Anything else to say on Kaggle, here? We will probably talk more about the corpus we analyzed later.}

@@

\AT{So, we understand that we iterated on our type system design.
  Subsequently, the analysis has undergone many iterations, but essentially
  it collects the most detailed information possible.  Do we want to mention
  all of that information, and how we ignore it later?  Or do we just want
  to focus on the relevant information for our final types?}

@@

%
%
%
%
%
%
\section{Type System Design}
\label{sec:typesystemdesign}

Our corpora analysis reported \AT{number} call traces for over \AT{number}
functions.  We distilled these call traces into function types which
captured the dynamic behaviour of the functions.  We emphasize that this was
an iterative process: initial analysis suggested a certain set of types,
which we tried and implemented, and further analysis suggested improvements,
and so on.  In this section, we will present the final type system that we
designed, which we believe captures the essence of R.  Throughout, we will
touch on design decisions and weigh our choices against alternatives,
speaking to the strengths and weaknesses of our chosen design.

\subsection{Type Grammar}
\label{subsec:typegrammar}

@@

A common primitive usage pattern is related to the presence of NAs:
\AT{often}, programmers explicitly check for NAs in primitive vectors,
possibly sanitizing them if they are present.  For instance, consider the
code in \AT{do an NA example}.  NAs complicate practically every computation
using primitive vectors, a fact that many programmers appear to be acutely
aware of.

@@

>>>>

\lstset{language=R}
\definecolor{LightGray}{rgb}{.92,.92,.92}
\definecolor{Gray}{rgb}{.3,.3,.3}
\definecolor{DarkGray}{rgb}{.5,.5,.5}
\lstset{ %
  columns=flexible,
  captionpos=b,
  frame=single,
  framerule=0pt,
  tabsize=2,
  belowskip=0.5em,
  backgroundcolor=\color{LightGray},
  basicstyle=\small\ttfamily,
  emphstyle=,
  keywordstyle=,
  commentstyle=\color{Gray}\em,
  stringstyle=\color{Gray},
  numbers=left
}
\lstdefinestyle{R}{ %
  language=R,
  morekeywords={assign, delayedAssign},
  deletekeywords={env, equal, c, runif, trace, args, exp, t, all},
  breaklines=true
}
\lstdefinestyle{Rin}{ %
  style=R,
  breaklines=false
}

@@

@@

@@

%
%
%
%
%
%
\section{Type Language Design}
\label{sec:typesystemdesign}

Our corpora analysis reported \AT{number} call traces for over \AT{number}
functions.  We distilled these call traces into function types which
captured the dynamic behaviour of the functions.  We emphasize that this was
an iterative process: initial analysis suggested a certain set of types,
which we tried and implemented, and reanalysis suggested improvements,
and so on.  In this section, we will present the final type annotation language design, 
which we believe captures the essence of R.  Throughout, we will
touch on design decisions and weigh our choices against alternatives,
speaking to the strengths and weaknesses of our chosen design.
We begin with basic, primitive types.

@@

A common primitive usage pattern is related to the presence of NAs:
\AT{often}, programmers explicitly check for NAs in primitive vectors,
possibly sanitizing them if they are present.  NAs complicate practically every computation
using primitive vectors, a fact that many programmers appear to be acutely
aware of. For instance, consider the code in Figure~\ref{fig:na-example}.

\begin{figure}[htbp]
\begin{center}

\begin{lstlisting}
binom.profile <- function(x, n, ...) {
  xn <- cbind(x = x, n = n)
  ok <- !is.na(xn[, 1]) & !is.na(xn[, 2])
  x <- xn[ok, "x"]
  n <- xn[ok, "n"]
  ...
\end{lstlisting}

\caption{NA checking example from the \code{binom} package.}
\label{fig:na-example}
\end{center}
\end{figure}

The \code{binom} package is a popular package \AT{do we want e.g. \# revdeps?} for computing confidence intervals for binomial experiments, and the \code{binom.profile} function does so using the profile likelihood, the details of which are not relevant here.
Here, we highlighted a relevant snippet of code, which appears at the beginning of the function: the programmer first binds the vectors into a matrix (line 2), with one column for each of \code{x} and \code{n}, then finds rows where both columns are not NA (line 3), then extracts non-NA values and stores them into \code{x} and \code{n} respectively (lines 4-5).
Importantly, this process is entirely quiet, and the user of \code{binom.profile} will not be notified if their data happens to contain NAs, and the NAs will be silently ignored. 
In reality, the presence of NAs in data could signify some sort of data corruption, and quietly ignoring issues in data could lead users (read: statisticians) to draw flawed conclusions from erroneous data.

To combat this, we introduce a modifier on primitive types: the type \code{^T} signifies that primitive type \code{T} might contain an NA value, while \code{T} signifies that \code{T} is NA-free. 
If the \code{binom} developers were to annotate arguments \code{x} and \code{n} with such an NA-free type, \contractr will notify users when \code{binom.profile} is called with NA-full values, rather than just quietly ignoring them.
\AT{Gather usage statistics.}

@@

%
%
%
%
\subsection{Final Type Grammar}
\label{subsec:typegrammar}
